Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-27 main.tex
fix

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!). So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

In this paper, we describe a compiler, Finch,
which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as
sparse tensors, databases, and lossless compression\kyle{nit: "unifying techniques in areas such as"}.
We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  
\end{abstract}

>>>>

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes. So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

In this paper, we describe a compiler, Finch,
which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
sparse tensors, databases, and lossless compression.
We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  
\end{abstract}

### 2024-03-27 sections/eval.tex
fix

The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.

Figures \ref{fig:spgemm_small} and \ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO. Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals. Additionally, TACO only supports outer product when the output matrix format is dense. It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices. In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.

As depicted in Figure \ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product. For large sparse matrices, as shown in Figure \ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05. It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.

@@

\begin{table}[htbp]
    \centering
    \caption{SpMV Sample Matrices}
    \label{spmv_sample_matrices}
    \begin{tabular}{|m{3cm}|c|c|c|}
        \hline
        \textbf{Spy} & \textbf{Group / Name} & \textbf{Dimensions} & \textbf{Best Finch Format}\\
        \hline
        \includegraphics[width=3cm]{spmv_matrices/saylr4.png} & HB/saylr4 & 3,564 x 3,564 (22,316) & Symmetric SparseList \\
        \hline 
        \includegraphics[width=3cm]{spmv_matrices/heart3.png} & Norris/heart3 & 2,339 x 2,339 (680,341) & SparseVBL \\
        \hline 
        \includegraphics[width=3cm]{spmv_matrices/toeplitz_large_band.png} & toeplitz\_large\_band  & 10,000 x 10,000 (1,999,900) & SparseBand \\
        \hline
        \includegraphics[width=3cm]{spmv_matrices/as-735.png} & SNAP/as-735 & 7,716 x 7,716 (26,467) & Symmetric SparseList-Pattern \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[htbp]
    \centering
    \caption{SpMV Sample Matrices}
    \label{spmv_sample_matrices}
    \begin{tabular}{|l|c|c|c|c|}
     \textbf{Spy} &
     \includegraphics[width=3cm]{spmv_matrices/saylr4.png} &
     \includegraphics[width=3cm]{spmv_matrices/heart3.png} &
     \includegraphics[width=3cm]{spmv_matrices/toeplitz_large_band.png} &

    \end{tabular}
\end{table}

@@

\begin{figure}
    \begin{minipage}[t]{0.5\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \includegraphics[width=\linewidth]{spmv_performance_sorted_(faster_than_taco).png}
    \end{minipage}%
    \begin{minipage}[t]{0.5\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \includegraphics[width=\linewidth]{spmv_performance_sorted_(slower_than_taco).png}
    \end{minipage}
    \caption{Performance of SpMV across various tools.}
    \label{spmv_sorted}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
    \caption{Performance of SpMV by Finch format.}
    \label{spmv_grouped}
    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. "R" indicates row-major implementation and "C" indicates column-major implementation in Finch. The baseline Finch format is unsymmetric Dense(SparseList(Element)).
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{erode2_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (2 iterations).}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (4 iterations).}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{hist_speedup_over_opencv.png}
    \caption{Performance of Finch on masked histogram task.}
\end{figure}

@@

To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path. We used graphs in GraphBLAS~\cite{mattson2019lagraph}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance. Direction-optimized BFS uses push and pull traversal to efficiently explore graphs. Push traversal starts from the source node, pushing neighboring nodes into the queue. Pull traversal starts from the target node, exploring predecessors (parents). This approach uses early-break in pull traversal, which is used to terminate the traversal prematurely once the target node is encountered, to avoid unnecessary edge traversal.

>>>>

The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.

Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.

Figures \ref{fig:spgemm_small} and \ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO. Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals. Additionally, TACO does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices. In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.

As depicted in Figure \ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product. For large sparse matrices, as shown in Figure \ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05. It's worth mentioning that TACO's Gustavson's algorithm is hard-wired.

@@

\begin{table}[htbp]
    \centering
    \caption{Transposed SpMV Sample Matrices}
    \label{tab:transposed_spmv_sample_matrices}
    \begin{tabular}{|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|}
        \hline
         & \textbf{HB/saylr4} & \textbf{Norris/heart3} & \textbf{large\_band} & \textbf{SNAP/as-735} \\
        \hline
        \textbf{Spy} & \includegraphics[width=\linewidth]{spmv_matrices/saylr4.png} & \includegraphics[width=\linewidth]{spmv_matrices/heart3.png} & \includegraphics[width=\linewidth]{spmv_matrices/toeplitz_large_band.png} & \includegraphics[width=\linewidth]{spmv_matrices/as-735.png} \\
        \hline 
        \textbf{Group / Name} & HB/saylr4 & Norris/heart3 & large\_band & SNAP/as-735 \\
        \hline
        \textbf{Dimensions} & 3,564 x 3,564 (22,316) & 2,339 x 2,339 (680,341) & 10,000 x 10,000 (1,999,900) & 7,716 x 7,716 (26,467) \\
        \hline
        \textbf{Best Finch Format} & Symmetric SparseList & SparseVBL & SparseBand & Symmetric SparseList-Pattern \\
        \hline
    \end{tabular}
\end{table}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
    \caption{Performance of SpMV by Finch format.}
    \label{fig:spmv_grouped}
    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. "R" indicates row-major implementation and "C" indicates column-major implementation in Finch. The baseline Finch format is unsymmetric Dense(SparseList(Element)).
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}
\end{figure}

%\begin{figure}
%	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
%    \caption{Performance of Finch on erosion task (4 iterations).}
%\end{figure}

@@

To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path. We used graphs in GraphBLAS~\cite{}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance. Direction-optimized BFS uses push and pull traversal to efficiently explore graphs. Push traversal starts from the source node, pushing neighboring nodes into the queue. Pull traversal starts from the target node, exploring predecessors (parents). This approach uses early-break in pull traversal, which is used to terminate the traversal prematurely once the target node is encountered, to avoid unnecessary edge traversal.
